---
title: "HDB Options for Singles 35+: A Net-Worth Decision Engine"
subtitle: "CDAR 2026 · Capstone Project Report"
author: "Team—Ivy · Lai Forng · Lee Yen"
date: last-modified
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
    number-sections: true
    fig-align: center
    fig-width: 9
    fig-height: 4.5
execute:
  warning: false
  message: false
  echo: true
editor: source
---

```{r}
#| label: setup
#| include: false
# Keep this lean—load only what the report actually uses.
library(tidyverse)   # dplyr / ggplot2 / readr / tidyr
library(tidymodels)  # rsample / recipes / parsnip / workflows / yardstick / tune
library(vip)         # variable importance for the M3 model
# library(sf)        # uncomment once you do the amenity -> town spatial join
# library(timetk)    # uncomment for M4 time_series_split
# library(modeltime) # uncomment for M4 model race
tidymodels_prefer()
theme_set(theme_minimal(base_size = 13))
set.seed(2026)
```

# Executive summary {#sec-exec}

**Decision (placeholder—replace with your computed result):** For a self-funded single aged 35+ earning *\$5,500/month*, the path that maximises projected *5-year net worth* is **[BTO / resale / rent]**, at roughly **\$[net-worth]** versus **\$[runner-up]** for the next-best path, under the central interest-rate scenario (**\~3.5%**) and central town-group price growth.

This report takes ONE HDB price dataset through Modules 1–4 and ends in a recommendation. The engine underneath is a **deterministic net-worth simulator**: given salary, age, retirement horizon and a rate scenario, it applies Singapore's financing rules (TDSR 55%, MSR 30%, LTV 75%, CPF for downpayment, lease decay) to compare BTO, resale and rent. Modules 2–4 supply the inputs the simulator needs: which **town group** you can afford (M2), a **price/psf model** to value a specific flat (M3), and a **directional 5-year growth range** per town group (M4).

**Two questions this answers**

- **Q-A (descriptive):** how do BTO, resale and rent compare *today* on upfront cash, monthly burden under the loan limits, and the town groups each opens up?
- **Q-B (prescriptive):** which path maximises projected **5-year net worth** under transparent rate scenarios and town-group growth ranges—and how sensitive is that to the assumptions?

**What we deliberately do NOT do:** we do not forecast 5-year interest rates (not credibly possible); we treat rates as **2–3 scenarios**. We do not run a causal Module 5—the question is a calculation, not an intervention (see @sec-m5).

```{r}
#| label: assumptions
#| eval: true
# Light, self-contained: the financing rules as one editable list.
# These are ILLUSTRATIVE defaults—TODO: confirm current MAS/HDB figures.
rules <- list(
  income_month   = 5500,     # stakeholder monthly income
  tdsr           = 0.55,     # total debt servicing ratio ceiling
  msr            = 0.30,     # mortgage servicing ratio ceiling (HDB)
  ltv            = 0.75,     # loan-to-value ceiling
  loan_years     = 25,       # tenure cap depends on age; TODO refine by age
  horizon_years  = 5         # net-worth evaluation window
)
# Rate SCENARIOS replace any attempt to forecast rates.
rate_scenarios <- tibble(
  scenario = c("HDB concessionary", "Bank low", "Bank high"),
  annual_rate = c(0.026, 0.035, 0.045)
)
rate_scenarios
```

# M1 · See it {#sec-m1}

**Goal:** build ONE clean panel—`town × flat_type × month` resale prices—and show the opening exhibit. This is the load-bearing module; everything downstream joins onto this panel.

```{r}
#| label: m1-import
#| eval: false
# flip to eval: true once the data.gov.sg CSVs are downloaded into ./data
# TODO: download from data.gov.sg and record URL + date in data/README.md
resale_raw <- read_csv("data/resale_flat_prices.csv")

resale <- resale_raw %>%
  mutate(
    month     = ym(month),                         # "2020-01" -> Date
    town      = str_to_title(town),
    flat_type = str_to_upper(flat_type),
    # remaining_lease often "61 years 04 months" -> numeric years
    remaining_lease_yrs = parse_remaining_lease(remaining_lease),
    price_psf = resale_price / floor_area_sqm      # keep a per-sqm measure
  ) %>%
  filter(flat_type %in% c("2 ROOM", "3 ROOM", "4 ROOM", "5 ROOM")) %>%
  select(month, town, flat_type, floor_area_sqm,
         remaining_lease_yrs, resale_price, price_psf)

glimpse(resale)
```

```{r}
#| label: m1-helper
#| eval: true
# Small, self-contained helper used above—safe to render without data.
# Parses "61 years 04 months" (or "61 years") into numeric years.
parse_remaining_lease <- function(x) {
  yrs <- str_extract(x, "\\d+(?=\\s*year)")  %>% as.numeric()
  mos <- str_extract(x, "\\d+(?=\\s*month)") %>% as.numeric()
  coalesce(yrs, 0) + coalesce(mos, 0) / 12
}
parse_remaining_lease(c("61 years 04 months", "80 years"))
```

```{r}
#| label: m1-panel
#| eval: false
# flip to eval: true once resale is loaded
# The panel the rest of the report joins onto.
town_panel <- resale %>%
  group_by(town, flat_type, month) %>%
  summarise(
    med_price = median(resale_price),
    med_psf   = median(price_psf),
    n         = n(),
    .groups   = "drop"
  )

# One opening exhibit: median price/psf over time, 4-ROOM, faceted sample of towns.
town_panel %>%
  filter(flat_type == "4 ROOM",
         town %in% c("Ang Mo Kio", "Punggol", "Bukit Merah",
                     "Woodlands", "Queenstown", "Sengkang")) %>%
  ggplot(aes(month, med_psf, colour = town)) +
  geom_line(linewidth = 0.8) +
  labs(title = "4-ROOM resale price per sqm over time",
       x = NULL, y = "Median price per sqm (S$)", colour = "Town")
```

# M2 · Group it {#sec-m2}

**Goal:** cluster the \~26 towns into interpretable groups, name them, and join the label forward as an M3 feature. **(Clustering is Module 2—this is the correction of the swapped M2/M3 label from the proposal.)**

```{r}
#| label: m2-features
#| eval: false
# flip to eval: true once town_panel exists
# One row per town: price level, recent growth, and (optional) amenity signals.
town_features <- town_panel %>%
  filter(flat_type == "4 ROOM") %>%
  group_by(town) %>%
  arrange(month, .by_group = TRUE) %>%
  summarise(
    level_psf   = median(med_psf),
    growth_5y   = last(med_psf) / first(med_psf) - 1,  # crude level growth
    volatility  = sd(med_psf),
    .groups = "drop"
  )
# TODO (stretch): left_join a few amenity counts/distances per town here
#   (MRT distance, hawker count, school count) from ONE sf::st_join.

glimpse(town_features)
```

```{r}
#| label: m2-cluster
#| eval: false
# flip to eval: true once town_features exists
set.seed(2026)
clust_rec <- recipe(~ level_psf + growth_5y + volatility, data = town_features) %>%
  step_normalize(all_numeric_predictors())

clust_prepped <- prep(clust_rec) %>% bake(new_data = NULL)

km <- kmeans(clust_prepped, centers = 3, nstart = 25)

town_groups <- town_features %>%
  mutate(cluster = factor(km$cluster)) %>%
  # TODO: inspect centroids, then rename to interpretable labels
  mutate(town_group = recode(cluster,
    `1` = "High-Yield",
    `2` = "Affordable-Growth",
    `3` = "Stable-Stagnant"
  ))

town_groups %>% count(town_group)
```

```{r}
#| label: m2-join-forward
#| eval: false
# flip to eval: true—this label is the feature M3 will test.
resale_labelled <- resale %>%
  left_join(town_groups %>% select(town, town_group), by = "town")
```

# M3 · Predict it {#sec-m3}

**Goal (supporting role):** a held-out price/psf model, and the graded **WITH vs WITHOUT the M2 town-group feature** comparison. This values a specific flat inside the simulator—it is not the net-worth engine itself.

```{r}
#| label: m3-split
#| eval: false
# flip to eval: true once resale_labelled exists
set.seed(2026)
model_df <- resale_labelled %>%
  drop_na(price_psf, remaining_lease_yrs, town_group) %>%
  mutate(across(c(flat_type, town_group), as.factor))

split   <- initial_split(model_df, prop = 0.8, strata = price_psf)
train   <- training(split)
test    <- testing(split)
```

```{r}
#| label: m3-recipes
#| eval: false
# flip to eval: true—two recipes: WITHOUT vs WITH the M2 cluster label.
rec_without <- recipe(price_psf ~ flat_type + remaining_lease_yrs + floor_area_sqm,
                      data = train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

rec_with <- recipe(price_psf ~ flat_type + remaining_lease_yrs + floor_area_sqm +
                     town_group, data = train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

rf_spec <- rand_forest(trees = 500) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("regression")

wf_without <- workflow() %>% add_recipe(rec_without) %>% add_model(rf_spec)
wf_with    <- workflow() %>% add_recipe(rec_with)    %>% add_model(rf_spec)
```

```{r}
#| label: m3-fit-score
#| eval: false
# flip to eval: true—HELD-OUT scoring on test, the number that counts.
set.seed(2026)
fit_without <- fit(wf_without, data = train)
fit_with    <- fit(wf_with,    data = train)

metric_set_reg <- metric_set(rmse, rsq, mae)

score <- function(fit_obj, label) {
  augment(fit_obj, new_data = test) %>%
    metric_set_reg(truth = price_psf, estimate = .pred) %>%
    mutate(model = label)
}

bind_rows(score(fit_without, "without M2 group"),
          score(fit_with,    "with M2 group")) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)
# TODO: state in prose whether the town_group feature earned its place.
```

```{r}
#| label: m3-vip
#| eval: false
# flip to eval: true—which drivers matter in the WITH model.
fit_with %>% extract_fit_parsnip() %>% vip(num_features = 10)
```

# M4 · Forecast it {#sec-m4}

**Goal (supporting role):** a **directional** 5-year growth range for each town GROUP (not per flat), with wide intervals. Monthly resale medians give enough points for `timetk`; annual series would be too small-n to trust, so we would say so and treat the forecast as directional only.

```{r}
#| label: m4-forecast
#| eval: false
# flip to eval: true once town_panel + town_groups exist
library(timetk); library(modeltime)

group_ts <- town_panel %>%
  filter(flat_type == "4 ROOM") %>%
  left_join(town_groups %>% select(town, town_group), by = "town") %>%
  group_by(town_group, month) %>%
  summarise(med_psf = median(med_psf), .groups = "drop")

# Model ONE group end-to-end; loop or nest over groups for the rest.
one_group <- group_ts %>% filter(town_group == "Affordable-Growth")

set.seed(2026)
splits <- time_series_split(one_group, date_var = month,
                            assess = "12 months", cumulative = TRUE)

m_arima <- arima_reg() %>% set_engine("auto_arima") %>%
  fit(med_psf ~ month, data = training(splits))
m_ets   <- exp_smoothing() %>% set_engine("ets") %>%
  fit(med_psf ~ month, data = training(splits))

mtbl <- modeltime_table(m_arima, m_ets)

mtbl %>%
  modeltime_calibrate(new_data = testing(splits)) %>%
  modeltime_accuracy()   # report MASE/RMSE; pick the winner

# TODO: refit winner on full history, forecast 60 months, keep WIDE intervals,
#   and feed the growth RANGE (not a point) into the simulator.
```

> **Note on rates:** we do NOT forecast interest rates. @sec-exec uses the three `rate_scenarios` instead. Two series trending together is not causation.

# M5 · Prescribe it {#sec-m5}

**Skipped, deliberately.** Module 5 answers "what happens if we *intervene*?" (A/B, DiD, synthetic control, matching). Our question is a **deterministic calculation** under fixed rules and scenarios—there is no treatment, no control, no counterfactual policy. Forcing a causal method here would be method without motive. The scenario grid in @sec-scenario is our sensitivity analysis instead.

# The net-worth simulator (the engine) {#sec-sim}

This deterministic function is the anchor deliverable: it applies the financing rules to each path and returns projected 5-year net worth. M2 sets the affordable town group, M3 values the flat, M4 supplies the growth range, scenarios set rates.

```{r}
#| label: simulator
#| eval: true
# Self-contained, illustrative: a transparent net-worth model, no external data.
# TODO: refine CPF logic, stamp duty, and rental assumptions with real figures.
monthly_payment <- function(principal, annual_rate, years) {
  r <- annual_rate / 12; n <- years * 12
  if (r == 0) principal / n else principal * r / (1 - (1 + r)^(-n))
}

simulate_networth <- function(path, price, annual_rate, growth,
                               rules, monthly_rent = 2200) {
  h  <- rules$horizon_years
  if (path == "rent") {
    # renter: no equity, pays rent; opportunity cost captured elsewhere (TODO)
    return(tibble(path = path, net_worth_5y = -monthly_rent * 12 * h))
  }
  loan   <- price * rules$ltv
  down   <- price - loan                              # from CPF/cash (TODO split)
  pmt    <- monthly_payment(loan, annual_rate, rules$loan_years)
  paid   <- pmt * 12 * h
  # crude amortised balance after h years:
  r <- annual_rate/12; n <- rules$loan_years*12; k <- h*12
  bal <- loan * ((1+r)^n - (1+r)^k) / ((1+r)^n - 1)
  value_5y <- price * (1 + growth)^h                  # M4 growth feeds here
  equity   <- value_5y - bal
  tibble(path = path, net_worth_5y = equity - paid - down)
}

# Illustrative call under the central scenario—replace price/growth with
# M3 valuation and M4 growth once available.
bind_rows(
  simulate_networth("bto",    price = 380000, annual_rate = 0.026,
                    growth = 0.03, rules = rules),
  simulate_networth("resale", price = 520000, annual_rate = 0.035,
                    growth = 0.02, rules = rules),
  simulate_networth("rent",   price = NA,     annual_rate = 0.035,
                    growth = 0,    rules = rules)
)
```

## Scenario-based analysis {#sec-scenario}

A 5-year interest-rate *forecast* is not credible, so we treat the two biggest unknowns—the **mortgage rate** and **town-group price growth**—as explicit **scenarios** and read the recommendation across all of them. This is the substitute for forecasting: rather than one fragile point prediction, we show *how the decision changes* as the assumptions move, and flag whether the winning path is **robust** (same winner everywhere) or **conditional** (the winner flips).

Two scenario axes give a 3 × 3 grid of futures for each path:

- **Rate**—HDB concessionary 2.6%, bank-low 3.5%, bank-high 4.5% (`rate_scenarios`).
- **Growth**—the 5-year price-growth range that M4 hands up, as low / central / high.

```{r}
#| label: growth-scenarios
#| eval: true
# The 5-year price-growth range from M4, expressed as three planning cases.
# TODO: replace with the low / central / high edges of your M4 forecast interval.
growth_scenarios <- tibble(
  growth_case = c("Low", "Central", "High"),
  growth      = c(0.005, 0.020, 0.035)   # annual price growth per town group
)
growth_scenarios
```

```{r}
#| label: scenario-sweep
#| eval: true
# Illustrative placeholder prices—TODO: replace with the M3 valuation for the
# specific flat, in the affordable town group from M2.
path_price <- c(bto = 380000, resale = 520000)

sweep <- tidyr::crossing(
  path = c("bto", "resale"),
  rate_scenarios,
  growth_scenarios
) %>%
  mutate(net_worth_5y = pmap_dbl(
    list(path, annual_rate, growth),
    ~ simulate_networth(..1, price = path_price[[..1]],
                        annual_rate = ..2, growth = ..3,
                        rules = rules)$net_worth_5y
  ))

sweep %>%
  transmute(path, scenario, rate = scales::percent(annual_rate, 0.1),
            growth_case, net_worth = scales::dollar(net_worth_5y)) %>%
  arrange(scenario, growth_case, path)
```

The **winner map** collapses each rate × growth cell to the better path and the margin—the single view a buyer reads:

```{r}
#| label: scenario-winner
#| eval: true
winner_map <- sweep %>%
  select(path, scenario, growth_case, net_worth_5y) %>%
  pivot_wider(names_from = path, values_from = net_worth_5y) %>%
  mutate(
    advantage_bto = bto - resale,                # > 0 => BTO wins the cell
    winner        = if_else(advantage_bto > 0, "BTO", "Resale"),
    margin        = scales::dollar(abs(advantage_bto))
  )

winner_map %>%
  transmute(scenario, growth_case, winner, margin,
            bto = scales::dollar(bto), resale = scales::dollar(resale))
```

```{r}
#| label: scenario-heatmap
#| eval: true
#| fig-cap: "BTO's net-worth advantage over resale across the scenario grid. Navy = BTO ahead, red = resale ahead; the pale band near zero is where the recommendation flips."
winner_map %>%
  mutate(
    scenario    = factor(scenario, levels = rate_scenarios$scenario),
    growth_case = factor(growth_case, levels = c("Low", "Central", "High"))
  ) %>%
  ggplot(aes(growth_case, scenario, fill = advantage_bto)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = paste0(winner, "\n", margin)), size = 3.3, lineheight = 0.9) +
  scale_fill_gradient2(low = "#b53a4a", mid = "#f3f6f9", high = "#1B3556",
                       midpoint = 0, labels = scales::dollar, name = "BTO − Resale") +
  labs(x = "Price-growth case (from M4)", y = "Rate scenario",
       title = "Where does the recommendation flip?") +
  theme(legend.position = "right")
```

**Reading it.** If every cell names the same path, the recommendation is **robust**—lead with it plainly. If the winner changes across cells, the decision is **conditional**: name the exact rate/growth combination where it flips and hand the buyer that trigger ("choose resale only if you expect growth above X *and* rates below Y").

**The flip point.** Holding the rate at the central 3.5% scenario, we sweep growth finely and solve for the break-even where the two paths draw level:

```{r}
#| label: scenario-flip
#| eval: true
flip <- tibble(growth = seq(0, 0.06, by = 0.0025)) %>%
  mutate(
    bto    = map_dbl(growth, ~ simulate_networth("bto",    path_price[["bto"]],
                     0.035, .x, rules)$net_worth_5y),
    resale = map_dbl(growth, ~ simulate_networth("resale", path_price[["resale"]],
                     0.035, .x, rules)$net_worth_5y),
    gap    = bto - resale
  )

flip_growth <- flip %>% slice_min(abs(gap), n = 1) %>% pull(growth)
cat(sprintf("At 3.5%% rates, BTO and resale break even near %.1f%% annual growth; below it BTO wins, above it resale pulls ahead.",
            100 * flip_growth))
```

**One-way sensitivity (tornado).** Which single assumption moves the answer most? From a base case (BTO, 3.5% rate, 2% growth) we push each lever to its scenario edges and record the swing in projected net worth—the widest bar is the number to research hardest before committing:

```{r}
#| label: scenario-tornado
#| eval: true
#| fig-cap: "How far each assumption alone moves BTO's projected 5-year net worth (low edge vs high edge). Dashed line = base case."
base <- list(price = path_price[["bto"]], rate = 0.035, growth = 0.02)
nw <- function(price, rate, growth)
  simulate_networth("bto", price, rate, growth, rules)$net_worth_5y

base_nw <- nw(base$price, base$rate, base$growth)

tornado <- tibble(
  lever = c("Mortgage rate", "Price growth", "Purchase price"),
  lo = c(nw(base$price, 0.045, base$growth),     # high rate -> lower net worth
         nw(base$price, base$rate, 0.005),
         nw(base$price * 1.10, base$rate, base$growth)),
  hi = c(nw(base$price, 0.026, base$growth),
         nw(base$price, base$rate, 0.035),
         nw(base$price * 0.90, base$rate, base$growth))
) %>%
  mutate(swing = abs(hi - lo)) %>%
  arrange(swing) %>%
  mutate(lever = factor(lever, levels = lever))

ggplot(tornado) +
  geom_segment(aes(y = lever, yend = lever, x = lo, xend = hi),
               linewidth = 7, colour = "#2B7A9B", lineend = "round") +
  geom_vline(xintercept = base_nw, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(labels = scales::dollar) +
  labs(x = "Projected 5-year net worth", y = NULL,
       title = "Sensitivity: which assumption matters most?",
       subtitle = "Base case: 3.5% rate, 2% growth")
```

::: {.callout-note title="Why this beats a forecast"}
A single rate forecast would be wrong with false precision. The scenario grid instead tells the buyer *what has to be true* for each path to win, and the tornado tells them *which number to research hardest*. That is the decision-useful output—and it is the sensitivity analysis that stands in for Module 5 here.
:::

# Recommendation {#sec-reco}

**Lead with the decision.** *(Fill from the computed sweep, not from intuition.)*

- **The call:** for a self-funded single aged 35+ on \$5,500/month, choose **[path]**—projected 5-year net worth **\$[net-worth]**, ahead of the next-best path by **\$[gap]** under the central rate scenario. *(see @sec-scenario)*
- **Q-A answered:** BTO needs the least upfront cash but has a wait; resale is available now at higher price/psf; renting builds no equity. *(see @sec-sim)*
- **Q-B answered:** the winning path **[does / does not]** flip across the rate and growth scenarios—the winner map and flip point in @sec-scenario show exactly where. The single assumption to pin down first is **[widest tornado bar]**.
- **Limits we name:** rates are scenarios, not forecasts; M4 growth is directional with wide intervals; the CPF/stamp-duty logic is simplified; the recommendation assumes the stated income is stable over 5 years.

```{r}
#| label: sessioninfo
#| eval: true
sessionInfo()
```
